Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-03-25 main.tex
updated section 6

\usepackage{cvpr}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\def\cvprPaperID{257} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

% Pages are numbered in submission mode, and unnumbered in camera-ready
\ifcvprfinal\pagestyle{empty}\fi
\begin{document}

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition}

@@

%With Hyperthreading enabled, each core can support two threads simultaneously.
For floating point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single
precision floating point values simultaneously.

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em vector-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.

@@

In summary, the alignment stage essentially two levels of available
parallelism: at the higher level there are multiple iterative alignent problems
that are solved independently, or {\bf problem-level} parallelism.  At a lower
level, the first-order linear algebraic operations exhibit parallelism within
image operations, i.e. at a pixel level.  We term this {\bf pixel-level}
parallelism.  In the following two sections will will discuss how to
exploit these two levels of parallelism on cpu and gpu architectures.

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}
This section discusses how to efficently map the two levels of parallelism
available in iterative face alignment to the two levels of concurrency
available in a typical multicore CPU architecture.

We will show that, in constrast with previous implementations, which mapped
problem level parallelism onto core-level concurrency, it is more efficient to
completely disregard problem-level parallelism, and instead map pixel-level
parallelism onto both core-level and vector-level cpu concurrency.  In other
words, the cpu implementation we advocate performs alignment of the test image
against the gallery images of a single user at a time.  Each matrix-vector and
vector-vector operation gets broken into 8 parts that run one per core, and
into 4 single float vectors that execute on the SSE SIMD units.
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_cpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_cpu}
\caption{Proposed mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_cpu}
\end{figure}

Recall that the L2 cache can old up to 128 image vectors' worth of data.  Since
we're only aligning data for one gallery user at a time the working set of the
problem fits into L2 cache.  Due to the higher bandwidth of L2 cache relative
to L3 cache, this results in significantly higher image alignment speed than
previous cpu implementations, as will be seen and discussed further in section
\ref{sec:alignment_benchmark}.

{\bf Implementation Details} In our implementation, most of the operations
required by the alignment stage take advantage of Intel Math Kernel Library
(MKL), a commercial implementation of the standard Basic Linear Algebra
Subprograms (BLAS).  For image resampling, we make use of Intel's Integrated
Performance Primitives (IPP) library.  Both of these libraries are heavily
optimized for Intel multicore CPU's, and are able to make good use of both
core-level and vector-level concurrency.  For operations that do not map well
onto existing libraries, we achieve both vector-level and core-level
concurrency through a combination of the OpenMP API \cite{dagum2002openmp} and
the automatic vectorization facilities of Intel's C++ compiler
(ICC)\cite{dulong1999overview}.

{\bf Comparison to Previous Implementations} The implementation presented in
\cite{WagnerA2009-CVPR} and \cite{WagnerA2011-PAMI} schedules instances of the
alignment problem to the cpu cores in a 1:1 fashion.  
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_old}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_old}
\caption{Old mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_old}
\end{figure}

This implementation was written using the OpenCV library
as a wrapper for MKL, and, like our proposed implementation, achieves vector-level concurrency implicitly.  
With eight alignment problems (associated with 8 training users) being solved
concurrently, and with 20-40 images per training user, all of the required data
fits into L3 cache.  While the L3 cache has much higher bandwidth and lower latency
than off-chip memory, this implementation is slower than the proposed implementation.

%During the course of development, many implementation strategies were tried.
%Listed in order of increasing performance based on a synthetic benchmark of
%the $\ell_1$ minimization step, they included:
%\begin{enumerate}
%\item A pure C++ implementation threaded and vectorized with OpenMP and icpc (200 ms).
%\item An OpenCV (with MKL backend) implementation threaded with OpenMP similar to \cite{WagnerA2011-PAMI} (190ms.)
%\item A straightforward MATLAB implementation (which uses threaded MKL internally) (60 ms.)
%\end{enumerate}

@@

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
This section discusses how to efficently map the two levels of parallelism
available in iterative face alignment to the two levels of concurrency
available in the NVIDIA Fermi GPU architecture.

Our proposed GPU implementation for the iterative alignment stage consists of a
single kernel that performs all of the test image alignment problems.  Each of
instance of the alignment problem is assigned its own thread block, and the GPU
hardware schedules as many of these thread blocks to run concurrently as the 
hardware will allow.  
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_gpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}
The full resolution test image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
recomputation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

\subsection{Other Implementation Strategies}
Several GPU implementation strategies were explored during the course of our
development, and are only mentioned briefly.  Listed in order of increasing
performance based on a synthetic benchmark of the $\ell_1$ minimization step,
they included:
\begin{enumerate}
\item Solving a single problem instance at a time using Cublas\footnote{Nvidia provided BLAS implementation} (302 ms)
\item Pure CUDA C implementation solving one problem per SMP using cuda streams\footnote{Streams is a CUDA feature that allows multiple
kernels to be run concurrently} (70 ms)
\item Pure CUDA C implementation solving one problem per SMP using one large kernel call (36 ms)
\end{enumerate}

@@

@@

\subsection{CPU Implementation} Our CPU implementation of Algorithm
\ref{alg:alm_rec} was written in C++.  For operations in the algorithm that map
nicely onto the BLAS API (such as SGEMV, SAXPY), we
call Intel's MKL BLAS library, which internally takes advantage of the SSE
units on all available CPU cores.

@@

The first experiment compares the performance of our two proposed implementations of 
the general $\ell_1$-minimization problem (Algorithm \eqref{alg:alm}) by
comparing the amount of time to solve a problem given a generic random matrix $A$ and a random sparse
vector $\xx_0$.  
The measurement $d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each entry
randomly generated from the standard normal and normalized with unit norm.  The ratio of $d:n$ varied from
$10:1$ to $10:5$ with $d$ varying from 500 to 8000.
The ground truth signals, $\xx_0$ had a sparsity rate of 0.1 with elements generated using the standard normal and normalized with unit norm.
Because the ground truth signal $\xx_0$ was known, the algorithm terminated
when $norm(\bx-\bx_0) < tolerance$.
The measurement vector was generated with the equation $\y_0 = A \x_0$.

>>>>

\usepackage{iccv}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\def\iccvPaperID{257} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

% Pages are numbered in submission mode, and unnumbered in camera-ready
\ificcvfinal\pagestyle{empty}\fi
\begin{document}

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}
}

@@

With Hyperthreading enabled, each core can support two threads simultaneously.
For floating point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single
precision floating point values simultaneously.

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em vector-level}
concurrency. CPU tools for exploiting these two levels of concurrency are
relatively mature.  Heavily optimized Basic Linear Algebra Subprograms (BLAS) are
commercially available that make good use of both levels of CPU concurrency.
For operations that do not fit well with the BLAS API, we 
achieve both vector-level and core-level concurrency through a combination of
the OpenMP API \cite{dagum2002openmp} and the automatic vectorization 
facilities of Intel's C++ compiler (ICC)\cite{dulong1999overview}.
% DW: I'm really not sure these citations are even really relevant;
% our target audience will just google icc openmp and get more up-to-date
% info anyway.
\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.

@@

In summary, the alignment stage essentially two levels of available
parallelism: at the higher level there are multiple iterative alignent problems
that are solved independently, or {\bf problem-level} parallelism.  At a lower
level, the first-order linear algebraic operations exhibit parallelism within
image operations, i.e. at a pixel level.  We term this {\bf pixel-level}
parallelism.

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}
One of the main contributions of this paper is a more efficent mapping of the
two levels of parallelism available in iterative face alignment to the two
levels of concurrency available in a typical multicore CPU architecture.

We argue that, in constrast with previous implementations, which mapped problem
level parallelism onto core-level concurrency, it is more efficient to
completely disregard problem-level parallelism, and instead map pixel-level
parallelism onto both core-level and vector-level cpu concurrency.  In other words,
the implementation we advocate performs alignment of the test image against the gallery
images of a single user at a time.  Each matrix-vector and vector-vector operation
gets broken into 8 parts that run one per core, and into 4 single float vectors
that execute on the SSE SIMD units.

Why is this
implementation faster?  Recall that the L2 cache can old up to 128 image
vectors' worth of data.  Since we're only aligning data for one gallery user at a
time the working set of the problem fits into L2 cache, resulting in
significantly higher image alignment speed than previous cpu implementations.

In our proposed CPU implementation of the iterative alignment stage, we
therefore perform parallelization in a manner similar to the matlab code:
solving a single problem instance at a time, relying on the threaded version of
MKL BLAS for both multi-core and vector parallelism.

This section discusses how the iterative alignment problem can be implemented
on CPU hardware.

{\bf Comparison to Previous Implementations}
The implementations presented in \cite{WagnerA2009-CVPR} and
\cite{WagnerA2011-PAMI} were written using the OpenCV library.  OpenCV wraps
either the operating system's built-in BLAS library, or Math Kernel Library
(MKL), Intel's BLAS implementation, and thus achieves vector (i.e. pixel)-level
parallelism implicitly.  The authors of \cite{WagnerA2011-PAMI} used OpenMP to
schedule instances of the alignment problem to the CPU cores in a 1:1 fashion.
With eight alignment problems for ten training users being solved concurrently,
and with 20-40 images per training user, all of the required data fits into L3
cache.  This provided a significant, though sub-linear improvement over a
single-core implementation.

During the course of development, many implementation strategies were tried.
Listed in order of increasing performance based on a synthetic benchmark of
the $\ell_1$ minimization step, they included:
\begin{enumerate}
\item A pure C++ implementation threaded and vectorized with OpenMP and icpc (200 ms).
\item An OpenCV (with MKL backend) implementation threaded with OpenMP similar to \cite{WagnerA2011-PAMI} (190ms.)
\item A straightforward MATLAB implementation (which uses threaded MKL internally) (60 ms.)
\end{enumerate}

@@

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}

This section discusses how to implement algorithm
\eqref{alg:iterative_alignment} on the GPU.  Like for the CPU case, many GPU
implementation strategies were explored during the course of our development,
and are only mentioned briefly.  Listed in order of increasing performance
based on a synthetic benchmark of the $\ell_1$ minimization step, they
included:
\begin{enumerate}
\item Solving a single problem instance at a time using Cublas\footnote{Nvidia provided BLAS implementation} (302 ms)
\item Pure CUDA C implementation solving one problem per SMP using cuda streams\footnote{Streams is a CUDA feature that allows multiple
kernels to be run concurrently} (70 ms)
\item Pure CUDA C implementation solving one problem per SMP using one large kernel call (36 ms)
\end{enumerate}

@@

Our proposed GPU implementation for the iterative alignment stage consists of a
single kernel that performs all of the test image alignment problems.  Each of
instance of the alignment problem is assigned its own thread block, and the GPU
hardware schedules as many of these thread blocks to run concurrently as the 
hardware will allow.

The full resolution test image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
recomputation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

@@

\subsection{CPU Implementation} Our CPU implementation of Algorithm
\ref{alg:alm_rec} was written in C++.  For operations in the algorithm that map
nicely onto the BLAS API (such as SGEMV, SAXPY), we
call Intel's MKL BLAS library, which internally takes advantage of the SSE
units on all available CPU cores. 
{\bf TODO: Our current implementation links against the MKL
library provided internally with matlab; we are in the process of updating the code to link directly against a newer version
of MKL}

@@

The first experiment compares the performance of our two proposed implementations of 
the general $\ell_1$-minimization problem (Algorithm \eqref{alg:alm}) given a generic random matrix $A$ and a random sparse
vector $\xx_0$.  
The measurement $d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each entry
randomly generated from the standard normal and normalized with unit column norm.  The ratio of $d:n$ varied from
$1:10$ to $5:10$ with $n$ varying from 500 to 8000.
The ground truth signals, $\xx_0$ had a sparsity rate of 0.1 with elements generated using the standard normal and also normalized with unit norm.
Because the ground truth signal $\xx_0$ was known, the algorithm terminated
when $\|\bx-\bx_0\| < \tau$.
The measurement vector was generated with the equation $\b_0 = A \x_0$.
